Supplemental Information B: Full Annotated Analysis and Reproducible Code

### 2021-10-20 Supplemental-Material-B-Annotated-Analysis.Rmd
AOT-S annotations and interpretations added

We start by creating an empty random intercept model (i.e., let the outgroup attitude intercept be different between participants; unconditional model)

@@

Comapring the two empty model, we find that there is indeed a significant amount of variance explained by including a random intercept.

To assess the variances within and between participants we look at the ICC $\tau_{00}^2 / (\tau_{00}^2 + \sigma^2)$: The ratio of the between-cluster variance to the total variance is called the Intraclass Correlation. It tells you the proportion of the total variance in Y that is accounted for by the clustering. (In this case the clustering means clustering observations per participant)

@@

In a final step we check whether the effect outgroup interactions, in part, depends on the quality during the interactions. Because we can only assess interaction quality when there is an interaction, it is difficult to assess this with the interaction dummy as a within person predictor. Instead, we will use an aggregate measure of interaction qualtity and average interaction quality to consider the two predictors jointly.

@@

Within the data, we find a medium sized correlation between the participants' Average Interaction Quality and their Average Outgroup Attitudes. Thus within our data participants with a higher quality outgroup interactions also held more positive attitudes towards that group. However, the frequency of intergroup interactions had no meaningfull correlation with either the average interaction quality or average outgroup attitudes.

@@

Given that we now have the perceived interaction quality as our outcome variable of interest we again begin with an unconditional model (i.e., empty random intercept model), to see whether there is enough variance to explain within the participants. Similarly to before this is again done wihin the subsample of measurements during which an outgroup interaction was reported.

@@

We add the core need fulfillment with the three self determiantion needs to a random intercept model of outgroup attitudes.

@@

@@

@@

@@

TL;DR: Neither the number of interactions nor the number of measurement beeps with an interaction are significantly related with the average outgroup attitudes. This is to say that within our data participants with more outgroup interactions did not have significantly more positive outgroup attitudes. This might be due to the aggregation within the participants or the small sample size of between participant data. Nonetheless, the aggregate data does not support the notion that simply having more interactions with an outgroup results in more positive outgroup attitudes.

@@

<div class="alert alert-info alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Note: </b><br/>
  mean difference alone probably even more stupid than only ignoring nested data, because we drop lots of data unnecessarily.  
  And only outgroup interaction Yes / No also weird because interaction group includes people who had only an outgroup interaction and those who also had an interaction with a non-outgroup person, and the no outgroup interaction group includes those who had no interaction at all and those who had an interaction with a non-outgroup person (but no outgroup interaction).
</div>

#### Regression with interaction term

Across participants: impact of different interaction types [probably meaningless because ignores response sets between participants]

@@

TL;DR: While controlling for interactions with non-Dutch people, outgroup attitudes were significantly higher when participants had an interaction with a Dutch person. The effect is relatively small (`r format(round(mdlStudent$lmAttInt$coefficients["OutgroupInteractionYes"], 2), nsmall=2)` points on a 0--100 scale). More importantly, however, this analysis lumps all ESM beeps from every participants together and ignores that the data is nested within participants.

@@

make a model with only the data but a random intercept (unconditional model)

@@

ICC $\tau_{00}^2 / (\tau_{00}^2 + \sigma^2)$: The ratio of the between-cluster variance to the total variance is called the Intraclass Correlation. It tells you the proportion of the total variance in Y that is accounted for by the clustering. (In this case the clustering means clustering observations per participant)

compare the random intercept model to a model without a random intercept (i.e., without levels at all)

@@

@@

tl;dr: Interaction with Dutch is great predictor, interactions with non-Dutch is not.

@@

@@

@@

TL;DR: There is no meaningfull correlation between the participants' Average Interaction Quality and their Average Outgroup Attitudes. However, the frequency of intergroup interactions had a significant correlation with both the average interaction quality and average outgroup attitudes.

@@

TL;DR: We find that, although the effect goes into the right direction (i.e., higher average interaction quality increases the positive effect of intergroup contact), there was no effect of average interaction quality (neither directly nor in a moderating role of interaction frequency). However, this regression ignores within person relationships of these variables (i.e., summarizes all within person variances into average scores).

@@

make a model with only the data but a random intercept (unconditional model) for the outgroup interaction subset.

@@

#### random intercept

@@

TL;DR: Situational core need relates significantly to outgroup attitudes. The core need has little explained variance between participants (compared to the null model: Variance Explained = 1 – (Var with Predictor/Var without Predictor); `r format(round(mdlStudentOut$Att$varBtwPercInterceptCore, 2),nsmall=2)`%). The variance explained within participants is small to medium (`r format(round(mdlStudentOut$Att$varWithinPercInterceptCore, 2),nsmall=2)`%)

@@

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Check needed: </b><br/>
  Something is with the variances explained (between = `r format(round(mdlStudentOut$Att$varBtwPercSlopesCore, 2),nsmall=2)`%, within = `r format(round(mdlStudentOut$Att$varWithinPercSlopesCore, 2),nsmall=2)`%). This is when comparing the random intercepts to the random slopes model.
</div>

@@

@@

make a model with only the data but a random intercept (unconditional model) for the outgroup interaction subset.

@@

TL;DR: Situational core need relates significantly to interaction quality. The core need explained substantial variance between participants (`r format(round(mdlStudentOut$Qlt$varBtwPercInterceptCore, 2),nsmall=2)`%). The variance explained within participants is also medium (`r format(round(mdlStudentOut$Qlt$varWithinPercInterceptCore, 2),nsmall=2)`%)

@@

TL;DR: random slopes does add significantly.

@@

TL;DR: Variance explained in outgroup attitudes is of medium effect size (between: `r format(round(mdlStudentOut$Att$varBtwPercInterceptCoreQlt,2),nsmall=2)`%, within: `r format(round(mdlStudentOut$Att$varWithinPercInterceptCoreQlt,2),nsmall=2)`%)

@@

#### Compare models

Interpretation of Mediation like results.

@@

@@

@@

@@

@@

# Compare new model to previous step
anova(
  mdlStudentOut$Att$lmeInterceptSdt, 
  mdlStudentOut$Att$lmeInterceptCoreSdt
  ) %>%
  as.data.frame() %>%
  select(-call) %>%
  mutate(
    L.Ratio = round(L.Ratio, 3),
    `p-value` = ifelse(`p-value`>=.001, round(`p-value`, 3), "< .001")
  ) %>%
  replace(is.na(.), "") %>%
  kbl(
    .,
    caption = "Student: Model Comparison",
    format = "html",
    linesep = "",
    booktabs = T,
    align = rep("c", ncol(.)),
    digits = 3
  ) %>%
  kable_styling(position = "left")

@@

@@

>>>>

We start by creating an empty random intercept model (i.e., let the outgroup attitude intercept be different between participants; unconditional model).

@@

Comparing the two empty model, we find that there is indeed a significant amount of variance explained by including a random intercept.

To assess the variances within and between participants we look at the ICC $\tau_{00}^2 / (\tau_{00}^2 + \sigma^2)$: The ratio of the between-cluster variance to the total variance is called the Intraclass Correlation. It tells you the proportion of the total variance in Y that is accounted for by the clustering. (In this case the clustering means clustering observations per participant).

@@

In a final step we check whether the effect outgroup interactions, in part, depends on the quality during the interactions. Because we can only assess interaction quality when there is an interaction, it is difficult to assess this with the interaction dummy as a within person predictor. Instead, we will use an aggregate measure of interaction quality and average interaction quality to consider the two predictors jointly.

@@

Within the data, we find a medium sized correlation between the participants' Average Interaction Quality and their Average Outgroup Attitudes. Thus within our data participants with a higher quality outgroup interactions also held more positive attitudes towards that group. However, the frequency of intergroup interactions had no meaningful correlation with either the average interaction quality or average outgroup attitudes.

@@

Given that we now have the perceived interaction quality as our outcome variable of interest we again begin with an unconditional model (i.e., empty random intercept model), to see whether there is enough variance to explain within the participants. Similarly to before this is again done within the subsample of measurements during which an outgroup interaction was reported.

@@

We add the core need fulfillment with the three self determination needs to a random intercept model of outgroup attitudes.

@@

The second sample we assess is larger study with international psychology students at Western European university. The hypotheses mirror those of the first study and are re-iterated here:

We will test our main hypotheses for this study in a sequential manner.

@@

We again test the most general contact hypothesis in two steps. First, we assess whether more intergroup interactions are related to to more positive outgroup attitudes. Second, we test whether a potential positive effect on outgroup attitudes depends on the interaction quality (jointly with the number of interactions).

@@

To test the impact of the overall number of interactions, we hope to find a significant relationship between the number of interactions a participant had and the average outgroup attitude.

@@

We find that both the number of interactions and the number of measurement beeps with an interaction are significantly related with the average outgroup attitudes. This is to say that within our data, participants with more outgroup interactions did have significantly more positive outgroup attitudes. This is inconsistent with the results we found in the worker sample.

@@

@@

We find that while controlling for interactions with non-Dutch people, outgroup attitudes were significantly higher when participants had an interaction with a Dutch person. The effect is of a medium size (`r format(round(mdlStudent$lmAttInt$coefficients["OutgroupInteractionYes"], 2), nsmall=2)` points on a 0--100 scale). However, this analysis lumps all ESM beeps from every participants together and ignores that the data is nested within participants.

@@

We start by creating an empty random intercept model (i.e., let the outgroup attitude intercept be different between participants; unconditional model).

@@

@@

Comparing the two empty model, we find that there is indeed a significant amount of variance explained by including a random intercept.

To assess the variances within and between participants we look at the ICC $\tau_{00}^2 / (\tau_{00}^2 + \sigma^2)$: The ratio of the between-cluster variance to the total variance is called the Intraclass Correlation. It tells you the proportion of the total variance in Y that is accounted for by the clustering. (In this case the clustering means clustering observations per participant).

@@

@@

We find that adding the random slopes does add significantly beyond the random intercept model. This is is different to the previous study where the random slope did not add significantly.

TL;DR: Random slopes adds significantly.

@@

In a final step we check whether the effect outgroup interactions, in part, depends on the quality during the interactions. Because we can only assess interaction quality when there is an interaction, it is difficult to assess this with the interaction dummy as a within person predictor. Instead, we will use an aggregate measure of interaction quality and average interaction quality to consider the two predictors jointly.

@@

Within the data, we find no significant correlation between the participants' Average Interaction Quality and their Average Outgroup Attitudes. Thus, within our data participants with a higher quality outgroup interactions did not hold more positive attitudes towards that group. However, the frequency of intergroup interactions had a meaningful correlation with both the average interaction quality or average outgroup attitudes.

@@

We find that in our student sample there is only a relationship between the number of outgroup contacts but no significant effect of average perceived contact quality. Nor do we find that in this sample the impact of the number of interactions is moderated by the average contact quality. This is not entirely consistent with the sojourner sample, where average contact quality did have a meaningful effect on outgroup attitudes. This effect is not necessarily surprising given that the variables aggregate all within person variation and there were substantially more measurements where participants did not have an interaction (but reported their outgroup attitudes) than measurements that followed an outgroup contact.

@@

@@

@@

We find that the the model with the added predictor indeed explains more variance in outgroup attitudes than the unconditional model. Looking at the individual coefficients, we find that the situational core need relates significantly to outgroup attitudes. The core need has little explained variance between participants (compared to the null model: Variance Explained = 1 – (Var with Predictor/Var without Predictor); `r format(round(mdlStudentOut$Att$varBtwPercInterceptCore, 2),nsmall=2)`%). The variance explained within participants is small to medium (`r format(round(mdlStudentOut$Att$varWithinPercInterceptCore, 2),nsmall=2)`%).

@@

<div class="alert alert-warning alert-dismissible fade in" role="alert">
  <a class="close" data-dismiss="alert" aria-label="close">&times;</a>
  <i class="fas fa-exclamation-triangle"></i> <b>Check needed: </b><br/>
  Something is with the variances explained (between = `r format(round(mdlStudentOut$Att$varBtwPercSlopesCore, 2),nsmall=2)`%, within = `r format(round(mdlStudentOut$Att$varWithinPercSlopesCore, 2),nsmall=2)`%). This is when comparing the random intercepts to the random slopes model. Check with materials from Laura!!
</div>

@@

Based on the assertion that the relationship between core need fulfillment and outgroup attitudes is explained by a higher perceived interaction, the core need fulfillment should also significantly predict perceived interaction quality.

@@

Given that we now have the perceived interaction quality as our outcome variable of interest we again begin with an unconditional model (i.e., empty random intercept model), to see whether there is enough variance to explain within the participants. Similarly to before this is again done within the subsample of measurements during which an outgroup interaction was reported.

@@

The predictor again adds a significant amount of explained variances beyond the empty model and looking at the slope coefficient, we find that the situational core need fulfillment relates significantly to perceived interaction quality. The core need explained substantial variance between participants (`r format(round(mdlStudentOut$Qlt$varBtwPercInterceptCore, 2),nsmall=2)`%). The variance explained within participants is also medium (`r format(round(mdlStudentOut$Qlt$varWithinPercInterceptCore, 2),nsmall=2)`%).

@@

We find that adding the random slopes does add significantly beyond the random intercept model.

@@

Unsurprisingly, the model with the two predictors adds significantly beyond the empty unconditional model. However, more importantly, looking at the coefficients, we find that the effect of core need fulfillment indeed is indeed strongly reduced and the variance is explained by the perceived interaction quality. The variance explained in outgroup attitudes is of medium effect size (between: `r format(round(mdlStudentOut$Att$varBtwPercInterceptCoreQlt,2),nsmall=2)`%, within: `r format(round(mdlStudentOut$Att$varWithinPercInterceptCoreQlt,2),nsmall=2)`%).

@@

@@

We find that the model explains significantly more variance than the empty null model. However, more interestingly, looking at the coefficients, we find that, as seen earlier, having an outgroup interaction has a strong effect on outgroup attitudes. Importantly, we find that there is a main effect of key need fulfillment but also a significant interaction effect of core need fulfillment and outgroup contact. This indicates that it is not simply key need fulfillment in general --- but especially key need fulfillment during an outgroup contact that predicts more positive outgroup attitudes.

@@

We find that adding the random slopes does add significantly beyond the random intercept model. We also see that when taking the possibility to varying slopes into account, the coefficient interpretations remains consistent (i.e., outgroup contact and its interaction with core need fulfillment remain important predictors of positive outgroup attitudes).

@@

The plots indicate that especially once we take the nested data structure into account we can see a substantially stronger effect of core need fulfillment on outgroup attitudes during outgroup contacts than without outgroup contacts.

@@

We add the core need fulfillment with the three self determination needs to a random intercept model of outgroup attitudes.

@@

# Compare new model to previous step
anova(
  mdlStudentOut$Att$lmeNull,
  mdlStudentOut$Att$lmeInterceptSdt, 
  mdlStudentOut$Att$lmeInterceptCoreSdt
  ) %>%
  as.data.frame() %>%
  select(-call) %>%
  mutate(
    L.Ratio = round(L.Ratio, 3),
    `p-value` = ifelse(`p-value`>=.001, round(`p-value`, 3), "< .001")
  ) %>%
  replace(is.na(.), "") %>%
  kbl(
    .,
    caption = "Student: Model Comparison",
    format = "html",
    linesep = "",
    booktabs = T,
    align = rep("c", ncol(.)),
    digits = 3
  ) %>%
  kable_styling(position = "left")

@@

We find that the the model with the added predictor indeed explains more variance in outgroup attitudes than the unconditional model and we find that adding the core need adds further explained variance --- beyond the self determination needs. Looking at the individual coefficients, we find that the situational core need relates significantly to outgroup attitudes, that it is a stronger predictor than any of the self determination theory needs and that it assumes some of the variance explained by the self determination theory needs (when compared to a model without the core need).

@@

We find that adding the random slopes does add significantly beyond the random intercept model. We also see that when taking the possibility to varying slopes into account, the coefficient interpretations remains consistent (i.e., core need and relatedness remain the strongest and only significant predictors).
